---
title: "R Tools for IMC Data"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('imcRtools')`"
author:
- name: Nils Eling
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
  email: nils.eling@dqbm.uzh.ch
- name: Tobias Hoch
  affiliation: 
  - Department for Quantitative Biomedicine, University of Zurich
  - Institute for Molecular Health Sciences, ETH Zurich
output:
    BiocStyle::html_document:
        toc_float: yes
bibliography: library.bib
abstract: |
    Abstract
vignette: |
    %\VignetteIndexEntry{"R Tools for IMC Data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE,
                        fig.retina = 0.75, crop = NULL)
library(BiocStyle)
```

```{r library, echo=FALSE}
library(imcRtools)
```

# Introduction

This vignette gives an introduction to .....

# Create a SingleCellExperiment Data Container

The `generateSCE` function allows to parse all relevant data and store in a handy `SingleCellExperiment` container. Find more information on the required input data in the help section. 

```{r quickstart-plotPixels}
#sce <- generateSCE(cells = cells,
#                   cell_meta = c("ImageNumber", "ObjectNumber"),
#                   image_meta = image_mat,
#                   panel_meta = panel_mat,
#                   metal_tag = "Metal.Tag",
#                   channel_number = "channel", 
#                   target_names = "clean_target",
#                   imageNumber = "ImageNumber",
#                   objectNumber = "ObjectNumber",
#                   measurement_column = "Intensity_MeanIntensityComp_FullStackFiltered",
#                   scaling_column = "Scaling_FullStack")
```

Once the data is loaded and stored, further `imcRtools` allow to visualize aspects of the data. 

# Acknowledgements


# Contributions


# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
