---
title: "Download example spillover data"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script downloads and re-formats example data for spillover correction.
As test data, we use the files that are provided by the following publication:

[Chevrier, Stéphane, Helena L. Crowell, Vito R. T. Zanotelli, Stefanie Engler, Mark D. Robinson, and Bernd Bodenmiller. 2017. “Compensation of Signal Spillover in Suspension and Imaging Mass Cytometry.” Cell Systems 6: 612–20.](https://www.cell.com/cell-systems/fulltext/S2405-4712(18)30063-2)

They can be accessed from [https://data.mendeley.com/datasets/v58yj49pfr/1](https://data.mendeley.com/datasets/v58yj49pfr/1).

More specificaly, the spillover files are part of `Figure S5`.

## Download the data

Here, we will first download the data.

```{r download-data}
sm_url <- "https://data.mendeley.com/public-files/datasets/v58yj49pfr/files/b39223d2-2825-4e79-9875-86fa0e1c55d2/file_downloaded?dl=1"

download.file(sm_url, "Figure_S5.zip")
unzip("Figure_S5.zip", overwrite=TRUE)
unlink("Figure_S5.zip")
```

## Read in and subset the data

We will now prepare a subsample of the data for testing purposes:

```{r subset-data, message=FALSE}
library(readr)
full_txt <- list.files("Figure_S5/Spillover_Matrix_1/", pattern = "Dy161|Dy162|Dy163|Dy164", full.names = TRUE)
full_txt_names <- list.files("Figure_S5/Spillover_Matrix_1/", pattern = "Dy161|Dy162|Dy163|Dy164")

full_txt <- lapply(full_txt, read_delim, delim = "\t")

## Subset to first 100 rows
full_txt <- lapply(full_txt, function(x){x[seq_len(100),]})

## Subset to specific channels
full_txt <- lapply(full_txt, function(x){x[,colnames(x) %in% 
                                               c("Start_push", "End_push", "Pushes_duration",
                                                 "X", "Y", "Z", "161Dy(Dy161Di)", 
                                                 "162Dy(Dy162Di)", "163Dy(Dy163Di)", "164Dy(Dy164Di)")]})
```

## Write out files

We will now save the processed files.

```{r save-files}
dir.create("../extdata/spillover")
for (i in seq_len(length(full_txt))) {
    write_delim(full_txt[[i]], 
                file = paste0("../extdata/spillover/", full_txt_names[i]),
                delim = "\t")
}
```

## Remove unnecessary files

Finally, we remove all unneeded files:

```{r remove-files}
unlink("Figure_S5/", recursive = TRUE)
```